% ============================================================
%  Vulnerability Type Classification with CodeBERT
%  ACL 2023 Format — compile with:
%    pdflatex main.tex && bibtex main && pdflatex main.tex x2
%
%  Style files: download acl.sty and acl_natbib.sty from
%  https://github.com/acl-org/acl-style-files
% ============================================================
\documentclass[11pt]{article}
\usepackage[]{acl}           % ACL 2023 style (place acl.sty in report/)
\usepackage{times}
\usepackage{latexsym}
\usepackage{microtype}
\usepackage{inconsolata}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{url}
\usepackage{xcolor}
\usepackage{listings}

% Code listing style
\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  backgroundcolor=\color{gray!8},
  keywordstyle=\color{blue!70},
  commentstyle=\color{green!50!black},
  stringstyle=\color{red!60},
}

\title{Vulnerability Type Classification in Source Code:\\
       A Comparative Study of TF-IDF Baselines and Fine-Tuned CodeBERT}

\author{
  \textbf{[Author Name(s)]} \\
  [Institution / Course Name] \\
  \texttt{[email@example.edu]}
}

\date{}

\begin{document}
\maketitle

% ============================================================
\begin{abstract}
Automated classification of source code vulnerabilities by weakness type
(CWE category) is a critical step toward scalable static analysis and
developer tooling. In this work we compare a traditional TF-IDF +
Logistic Regression baseline against a fine-tuned
\textsc{CodeBERT}~\citep{feng-etal-2020-codebert} model on the
BigVul dataset~\citep{fan2020ac}, which contains over 3,000 C/C++
vulnerability-fixing commits labeled with CWE identifiers. Our
CodeBERT model achieves an accuracy of \textbf{[X.XXXX]} and a
macro-averaged F1 score of \textbf{[X.XXXX]}, outperforming the
baseline by [X.XX] and [X.XX] points respectively. We further analyze
per-class performance and identify the confusion patterns most common
among structurally similar CWE categories.
\end{abstract}

% ============================================================
\section{Introduction}
\label{sec:intro}

Software vulnerabilities cost the global economy billions of dollars
annually~\citep{anderson2020economics}. The National Vulnerability
Database (NVD) catalogues thousands of new Common Vulnerabilities and
Exposures (CVEs) each year, each associated with one or more Common
Weakness Enumeration (CWE) identifiers that describe the \emph{type}
of weakness—buffer overflow, use-after-free, SQL injection, and so on.
Automatically assigning CWE labels to vulnerable code enables
developers to receive targeted remediation guidance and helps
prioritizers triage security debt at scale.

Classifying code by vulnerability type is harder than sentiment
analysis or topic modeling for several reasons: (1) the relevant
signal is often localized to a few lines within a function of
hundreds of lines; (2) different vulnerability types can produce
syntactically similar constructs; and (3) training data is scarce
relative to the diversity of CWE categories.

Transformer models pre-trained on source code—most notably
\textsc{CodeBERT}~\citep{feng-etal-2020-codebert}—have demonstrated
strong transfer learning capabilities on downstream code-understanding
tasks such as clone detection, defect detection, and code search.
We hypothesize that these models can similarly benefit vulnerability
type classification. To test this hypothesis we formulate the problem
as a multi-class text classification task and conduct the following
study:

\begin{enumerate}
  \item Train a TF-IDF character n-gram + Logistic Regression baseline.
  \item Fine-tune \textsc{CodeBERT} on the same labeled splits.
  \item Evaluate both models with accuracy, macro-F1, precision, and
        recall on a held-out test set.
  \item Perform per-class and error analysis to understand failure
        modes.
\end{enumerate}

The remainder of this paper is structured as follows.
Section~\ref{sec:related} reviews related work.
Section~\ref{sec:data} describes the dataset.
Section~\ref{sec:method} describes the models.
Section~\ref{sec:experiments} details experimental settings.
Section~\ref{sec:results} presents results.
Section~\ref{sec:discussion} discusses findings.
Section~\ref{sec:conclusion} concludes.

% ============================================================
\section{Related Work}
\label{sec:related}

\paragraph{Vulnerability Detection.}
Early work on automated vulnerability detection relied on hand-crafted
features and pattern matching~\citep{chess2004static}. Deep learning
approaches using bidirectional LSTMs and graph neural networks on
code property graphs achieved notable performance on binary
(vulnerable / not-vulnerable) detection~\citep{li2018vuldeepecker,
zhou2019devign}. Our task is harder: we require \emph{multi-class}
CWE labeling, not just binary detection.

\paragraph{Pre-trained Models for Code.}
\textsc{CodeBERT}~\citep{feng-etal-2020-codebert} is a bimodal model
pre-trained on natural language–code pairs from GitHub using masked
language modeling (MLM) and replaced token detection (RTD) objectives.
It achieves state-of-the-art results on code search, code summarization,
and defect detection. \textsc{GraphCodeBERT}~\citep{guo2021graphcodebert}
extends this with data flow graphs, while \textsc{UniXcoder}~\citep{guo2022unixcoder}
unifies encoder and decoder pre-training. We choose \textsc{CodeBERT}
as it balances capability with training efficiency.

\paragraph{CWE Classification.}
\citet{aota2021automation} used TF-IDF and SVMs on CVE descriptions
(natural language) rather than source code. \citet{ni2022just}
applied CodeBERT to a binary defect detection task but did not perform
fine-grained CWE type classification. To our knowledge, this work is
among the first to directly fine-tune \textsc{CodeBERT} for multi-class
CWE labeling on raw function-level source code.

% ============================================================
\section{Dataset}
\label{sec:data}

\subsection{BigVul}

We use the BigVul dataset~\citep{fan2020ac}, a collection of
186,878 C/C++ functions extracted from [X,XXX] vulnerability-fixing
commits across [XXX] open-source projects. Each function is paired
with a CVE identifier and—for vulnerable functions—one or more CWE
labels. We retain only the top-10 most frequent CWE categories to
avoid extreme class imbalance, yielding [X,XXX] labeled samples.

\subsection{Preprocessing and Splits}

We apply minimal preprocessing: inline and block comments are removed,
and whitespace is normalized. We perform a stratified 70 / 15 / 15
train / validation / test split, ensuring each CWE class is
proportionally represented in every partition.

Table~\ref{tab:data} summarizes the final dataset statistics.

\begin{table}[h]
\centering
\small
\begin{tabular}{lrr}
\toprule
\textbf{Split} & \textbf{Samples} & \textbf{Classes} \\
\midrule
Train      & [X,XXX] & 10 \\
Validation & [X,XXX] & 10 \\
Test       & [X,XXX] & 10 \\
\midrule
Total      & [X,XXX] & 10 \\
\bottomrule
\end{tabular}
\caption{Dataset splits after filtering to top-10 CWE categories.}
\label{tab:data}
\end{table}

\subsection{Class Distribution}
[FILL IN: describe the most frequent CWEs in your filtered set, e.g.,
CWE-119 (Buffer Errors), CWE-20 (Input Validation), CWE-200
(Information Exposure). Note any imbalance and whether you applied
class weighting.]

% ============================================================
\section{Methodology}
\label{sec:method}

\subsection{Baseline: TF-IDF + Logistic Regression}

Our baseline uses a \textbf{character-level TF-IDF} vectorizer
($n$-grams of length 3–6) with 100,000 features combined with
\textbf{Logistic Regression} (multi-class ``multinomial'',
$C{=}1.0$, balanced class weights, L-BFGS solver, 1,000 iterations).
Character n-grams capture syntactic patterns in source code such as
operator sequences and identifier suffixes without requiring
tokenization.

\subsection{CodeBERT Fine-Tuning}

We add a linear classification head over the \texttt{[CLS]} token
representation of \texttt{microsoft/codebert-base} (125M parameters,
12 transformer layers, hidden size 768):

\begin{equation}
  \hat{y} = \text{softmax}(W \cdot h_{\texttt{[CLS]}} + b),
  \quad W \in \mathbb{R}^{K \times 768}
\end{equation}

where $K$ is the number of CWE classes. We fine-tune all parameters
jointly using cross-entropy loss. Input functions are tokenized with
the \texttt{RobertaTokenizer} and truncated / padded to 512 tokens.

% ============================================================
\section{Experiments}
\label{sec:experiments}

\subsection{Implementation Details}

All experiments are run on a single NVIDIA [GPU MODEL] GPU with
[X]GB of VRAM. PyTorch [VERSION] and Hugging Face \texttt{transformers}
[VERSION] are used throughout.

\paragraph{Baseline.} The scikit-learn pipeline fits in under 5 minutes
on CPU.

\paragraph{CodeBERT.}
\begin{itemize}
  \item Optimizer: AdamW ($\beta_1{=}0.9$, $\beta_2{=}0.999$,
        $\epsilon{=}10^{-8}$)
  \item Learning rate: $2 \times 10^{-5}$ with linear warmup over
        10\% of total steps, then linear decay to 0
  \item Weight decay: $0.01$
  \item Batch size: 16
  \item Epochs: 5
  \item Gradient clipping: $\ell_2$ norm $\leq 1.0$
  \item Dropout: 0.1 (hidden and attention)
  \item Best checkpoint selected by macro-F1 on validation set
\end{itemize}

\subsection{Evaluation Metrics}

We report \textbf{accuracy}, \textbf{macro-averaged F1}, macro-averaged
\textbf{precision}, and macro-averaged \textbf{recall}. Macro averaging
weights all classes equally, which is appropriate given the class
imbalance in our filtered dataset.

% ============================================================
\section{Results}
\label{sec:results}

\subsection{Main Results}

Table~\ref{tab:results} compares the baseline and CodeBERT on the
held-out test set.

\begin{table}[h]
\centering
\small
\begin{tabular}{lcccc}
\toprule
\textbf{Model} & \textbf{Acc} & \textbf{F1} & \textbf{P} & \textbf{R} \\
\midrule
TF-IDF + LR  & [X.XXXX] & [X.XXXX] & [X.XXXX] & [X.XXXX] \\
CodeBERT     & [X.XXXX] & [X.XXXX] & [X.XXXX] & [X.XXXX] \\
\midrule
$\Delta$     & [+X.XX]  & [+X.XX]  & [+X.XX]  & [+X.XX]  \\
\bottomrule
\end{tabular}
\caption{Test-set performance. $\Delta$ = CodeBERT $-$ Baseline.
         All metrics are macro-averaged; higher is better.}
\label{tab:results}
\end{table}

\subsection{Per-Class Results}

Table~\ref{tab:perclass} shows per-class F1 for both models on the
test set.

\begin{table}[h]
\centering
\small
\begin{tabular}{lcc}
\toprule
\textbf{CWE} & \textbf{Baseline F1} & \textbf{CodeBERT F1} \\
\midrule
CWE-119 (Buffer Errors)      & [X.XXXX] & [X.XXXX] \\
CWE-20  (Input Validation)   & [X.XXXX] & [X.XXXX] \\
CWE-200 (Info Exposure)      & [X.XXXX] & [X.XXXX] \\
CWE-264 (Permissions)        & [X.XXXX] & [X.XXXX] \\
CWE-362 (Race Condition)     & [X.XXXX] & [X.XXXX] \\
CWE-416 (Use After Free)     & [X.XXXX] & [X.XXXX] \\
CWE-476 (NULL Deref)         & [X.XXXX] & [X.XXXX] \\
CWE-787 (Out-of-Bounds Write)& [X.XXXX] & [X.XXXX] \\
CWE-862 (Missing Auth.)      & [X.XXXX] & [X.XXXX] \\
CWE-125 (Out-of-Bounds Read) & [X.XXXX] & [X.XXXX] \\
\midrule
\textbf{Macro Avg}           & [X.XXXX] & [X.XXXX] \\
\bottomrule
\end{tabular}
\caption{Per-class F1 scores on the test set.
         [FILL IN with your actual CWE classes.]}
\label{tab:perclass}
\end{table}

% ============================================================
\section{Discussion}
\label{sec:discussion}

\subsection{Effect of Pre-Training}
[FILL IN: Discuss why CodeBERT outperforms the baseline. Reference
the contextualized token representations vs. bag-of-ngrams. Discuss
whether the improvement is larger for syntactically complex CWEs
(e.g., race conditions) vs. those with distinctive keywords
(e.g., CWE-89 SQL injection).]

\subsection{Confusion Patterns}
[FILL IN: Based on your confusion matrix, identify the most common
misclassification pairs. A common pattern is confusion between
CWE-119 and CWE-787 / CWE-125 because all involve memory boundary
violations. Discuss what features would be needed to separate them.]

\subsection{Qualitative Error Analysis}
[FILL IN: Describe 2–3 specific misclassified examples. What does
the model predict and why might it be confused? Are the functions
very short, multi-labeled in the original dataset, or domain-specific?]

\subsection{Limitations}
This study has several limitations. First, we restrict to top-10
CWE classes; the tail of the distribution remains unaddressed.
Second, truncating functions to 512 tokens discards context for
longer functions. Third, class imbalance, though partially addressed
via balanced weighting, still influences minority-class performance.
Future work could explore \textsc{LongFormer}-style models, graph-based
representations, or data augmentation strategies.

% ============================================================
\section{Conclusion}
\label{sec:conclusion}

We presented a systematic comparison of a TF-IDF baseline and a
fine-tuned \textsc{CodeBERT} model for multi-class CWE vulnerability
type classification. Our CodeBERT model achieves [X.XXXX] accuracy
and [X.XXXX] macro-F1 on the BigVul dataset, surpassing the baseline
by [X.XX] and [X.XX] points respectively. These results confirm that
pre-trained code representations transfer effectively to fine-grained
vulnerability categorization. Error analysis reveals that the most
challenging cases involve structurally similar memory-safety CWEs,
suggesting that incorporating data flow or control flow information
could further improve performance.

% ============================================================
\section*{Acknowledgements}
[FILL IN or remove this section.]

% ============================================================
\bibliography{references}

\end{document}
